---
title: "CARI console indicators"
output: html_document
date: "2023-01-05"
toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

```{r}
library(dplyr)
library(openxlsx)
library(broom)
library(tidyr)
library(stringr)
library(ggplot2)
library(xtable)
library(gtsummary)
library(fauxnaif)
library(geojsonio)
library(broom)
library(viridis)
library(mice)

```
# Dataset

```{r}
main_ds <- read.xlsx("UKR2308_MSNA_clean_data_230825_shared_no_pass.xlsx", sheet = 1)
hh_members <- read.xlsx("UKR2308_MSNA_clean_data_230825_shared_no_pass.xlsx", sheet = 2)

#main_ds_copy <- read.xlsx("C:\\Users\\hanna.kurovska\\OneDrive - World Food Programme\\Desktop\\RAM\\MSNA\\Dataset\\Clean\\REACH_WFP_MSNA_UKR_Final_Data_Internal.xlsx", sheet = "main")
```


mutate uniform oblast column for F2F and CATI data

```{r}
main_ds <- main_ds |> 
  mutate(oblast_total = coalesce(oblast, O_4_current_location_oblast))
```


# FCS

FCS variables

"Prefer not to answer" is turned to 0.

```{r}
nam_main_ds <- names(main_ds)
fcs_names <- nam_main_ds[grepl("fcs", nam_main_ds)]
# 2023
#fcs_names <- fcs_names[grepl("number", fcs_names)]

frt <- NA
for(f in fcs_names){
  t <- main_ds |> 
    count(!!sym(f)) |>
    mutate(prp = n/sum(n),
           var = f) |>
    setNames(c("cat", "n", "prp", "var"))
  frt <- rbind(frt, t)
}

 i <- 1
 for(i in 1:length(fcs_names)){
   
   main_ds[,paste0(fcs_names[i])] <- ifelse(main_ds[,paste0(fcs_names[i])] == "prefer_not_to_answer", "0", main_ds[,paste0(fcs_names[i])]) 
   
   main_ds[,paste0(fcs_names[i])] <- as.integer(main_ds[,paste0(fcs_names[i])])
 }
```

FCS calculation

```{r}
#main_ds <- main_ds |>
#  mutate(fcs = mapply(sum, (G_1_1_fcs_cereals*2), (G_1_2_fcs_legumens*3), (G_1_3_fcs_dairy*4), (G_1_4_fcs_meat*4), (G_1_5_fcs_vegetables), (G_1_6_fcs_fruits), (G_1_7_fcs_oil*0.5), (G_1_8_fcs_sugar*0.5)))

# 2023
main_ds <- main_ds |>
  mutate(across(all_of(fcs_names), as.numeric))

main_ds <- main_ds |>
  mutate(fcs = mapply(sum, (G_1_1_fcs_cereals*2), (G_1_2_fcs_legumens*3), (G_1_6_fcs_dairy*4), (G_1_5_fcs_meat*4), (G_1_3_fcs_vegetables), (G_1_4_fcs_fruits), (G_1_7_fcs_oil*0.5), (G_1_8_fcs_sugar*0.5)))
```

Removing zero FCS (replacing to NA)

```{r}
main_ds |> count(fcs < 7)

main_ds <- main_ds |>
  mutate(fcs = ifelse(fcs < 7, NA, fcs))
```


Plotting FCS

```{r}
ggplot(main_ds, aes(x = fcs))+
  geom_area(stat = "bin", alpha = 0.8)+
  theme_minimal()+
  geom_vline(xintercept = 28, color = "red")+
  geom_vline(xintercept = 42, color = "red")
```


FCS groups (Ukraine) (28-42 treshold)

```{r}
main_ds$fcs_groups <- cut(main_ds$fcs, breaks = c(0, 28, 42, Inf), include.lowest = T, labels = c("Poor", "Borderline", "Acceptable"))
```

Plot FCS groups

```{r}
ggplot(main_ds, aes(x = O_1_respondent_consent, fill = fcs_groups))+ # 2023 - variable change
  geom_bar(position = "fill", stat = "count", width = 0.5)+
  theme_minimal()+
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank()
  )
```
FCS groups frequencies

```{r}
#main_ds |> select(o1_current_oblast, fcs_groups) |> tbl_summary(by = o1_current_oblast) |> add_overall()
# 2023
main_ds |> select(oblast_total, fcs_groups) |> tbl_summary(by = oblast_total) |> add_overall()

#main_ds |> group_by(fcs_groups) |> summarise(n = n()) |> ungroup() |> mutate(prp = n/sum(n))
```

# rCSI

rCSi calculation

(Prefer not to answer is getting replaced by 0, and obs. that have all rCS variables "prefer not to answer", are replaced to NA)

```{r}
rcs_names <- nam_main_ds[grepl("rcs", nam_main_ds)]

# change: removing "all rCSi--variables are "prefer not to answer""
main_ds <- main_ds |> 
  mutate(rcs_to_NA = case_when(if_all(rcs_names, ~ . == "prefer_not_to_answer") ~ "rm", TRUE ~ "kp")) |>
  mutate(across(rcs_names, ~ case_when(rcs_to_NA == "kp" ~ .)))


main_ds <- main_ds |>
  mutate(across(rcs_names, str_replace, "prefer_not_to_answer", "0")) |> mutate(G_3_4_rcsi_restrict_consumption_adults = ifelse(is.na(G_3_4_rcsi_restrict_consumption_adults), "0", G_3_4_rcsi_restrict_consumption_adults)) |>
  mutate(across(rcs_names, as.numeric))

table(main_ds$rcs_to_NA)
```

Calculating rCSI

```{r}
main_ds <- main_ds |> mutate(rcsi = mapply(sum, (G_3_1_rcsi_cheaper_food), (G_3_2_rcsi_borrow_food*2), (G_3_5_rcsi_reduce_meals_number), (G_3_3_rcsi_limit_portion), (G_3_4_rcsi_restrict_consumption_adults*3)))
```


rCSI groups

```{r}
main_ds <- main_ds |> mutate(rcsi_groups = case_when(
  rcsi < 4 ~ "Low rCSI",
  rcsi >= 4 & rcsi <= 18 ~ "Medium rCSI",
  rcsi > 18 ~ "High rCSI"
))
```

Plotting rCSI groups

```{r}
ggplot(main_ds, aes(x = O_1_respondent_consent, fill = rcsi_groups))+
  geom_bar(position = "fill", stat = "count", width = 0.5)+
  theme_minimal()+
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank()
  )
```
rCSI groups frequencies

```{r}
main_ds |> select(oblast_total, rcsi_groups) |> tbl_summary(by = oblast_total) |> add_overall()
```


# Current status CARI domain
Groups for CARI (current status)

```{r}
main_ds$CARI_current_status <- case_when(
  main_ds$fcs_groups == "Acceptable" & main_ds$rcsi < 4 ~ 1,
  main_ds$fcs_groups == "Acceptable" & main_ds$rcsi >= 4 ~ 2,
  main_ds$fcs_groups == "Borderline" ~ 3,
  main_ds$fcs_groups == "Poor" ~ 4
)

```

Plotting a current status domain

```{r}
ggplot(main_ds, aes(x = O_1_respondent_consent, fill = as.factor(CARI_current_status)))+
  geom_bar(position = "fill", stat = "count", width = 0.5)+
  theme_minimal()+
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank()
  )
```

# LCS

Livelihood coping strategies

Mutating LCS (I_8_6 and I_8_7 are being merged together (reducing health or education expenditures))

```{r}
main_ds <- main_ds |> mutate(I_8_6_7_lcs_reduce_health_and_education_expenditures = case_when(
  I_8_6_lcs_reduce_health_expenditures == "yes" | I_8_7_lcs_reduce_education_expenditures == "yes" ~ "yes",
  I_8_6_lcs_reduce_health_expenditures == "no_have_already_exhausted_this_coping_strategy_and_cannot_use_it_again" | I_8_7_lcs_reduce_education_expenditures == "no_have_already_exhausted_this_coping_strategy_and_cannot_use_it_again" ~ "no_have_already_exhausted_this_coping_strategy_and_cannot_use_it_again",
  TRUE ~ "no"

))
```

Classifying LCS

  1. Stress: Spending savings, Borrowing food, Eating elsewhere, selling HH assets; -Taking on additional job (not present)
  2. crisis: Selling productive assets, Reducing health/education expenditures, Moving elsewhere
  3. Emergency: Using degrading income sources, Selling a house, Ask strangers for money

```{r}
lcs_vars <- names(main_ds)[str_detect(names(main_ds), "_lcs_")]
lcs_vars <- lcs_vars[! lcs_vars %in% c("I_8_6_lcs_reduce_health_expenditures", "I_8_7_lcs_reduce_education_expenditures", "I_8_12_lcs_move_another_dwelling", "I_8_13_lcs_abandon_hh_assets_goods", lcs_vars[str_starts(lcs_vars, "I_8_14_lcs_reason")])]


stress_vars <- c( "I_8_2_lcs_spent_savings", 
                  "I_8_3_lcs_forrowed_food", 
                  "I_8_4_lcs_eat_elsewhere", 
                  "I_8_1_lcs_sell_hh_assets")

crisis_vars <- c("I_8_5_lcs_sell_productive_assets", 
                 "I_8_6_7_lcs_reduce_health_and_education_expenditures", 
                 "I_8_9_lcs_move_elsewhere")

emergency_vars <- c("I_8_13_lcs_ask_stranger", 
                    "I_8_8_lcs_sell_house", 
                    "I_8_10_lcs_degrading_income_source")

main_ds <- main_ds |> 
  #rowwise() |> 
  mutate(
    NA_coping = ifelse(if_any(lcs_vars, ~ is.na(.)), 1, 0),
    no_coping = 
    ifelse(if_any(lcs_vars, ~ .  %in% c("yes", "no_have_already_exhausted_this_coping_strategy_and_cannot_use_it_again")), 0, 1),
    stress_coping = 
    ifelse(if_any(stress_vars, ~ .  %in% c("yes", "no_have_already_exhausted_this_coping_strategy_and_cannot_use_it_again")), 1, 0),
    crisis_coping = 
    ifelse(if_any(crisis_vars, ~ .  %in% c("yes", "no_have_already_exhausted_this_coping_strategy_and_cannot_use_it_again")), 1, 0),
    emergency_coping = ifelse(if_any(emergency_vars, ~ .  %in% c("yes", "no_have_already_exhausted_this_coping_strategy_and_cannot_use_it_again")), 1, 0))
```

```{r}
main_ds <- main_ds |> mutate(coping_strategy  = case_when(
  emergency_coping == 1 ~ "emergency",
  emergency_coping == 0 & crisis_coping == 1 ~ "crisis",
  emergency_coping == 0 & crisis_coping == 0 & stress_coping == 1 ~ "stress",
  emergency_coping == 0 & crisis_coping == 0 & stress_coping == 0 & NA_coping == 0 ~ "no_coping",
  #NA_coping == 1 ~ NA_character_,
  TRUE ~ NA_character_
))


```


```{r}

  main_ds <- main_ds |> mutate(lcs_cc = case_when(
    coping_strategy == "no_coping" ~ 1,
    coping_strategy == "stress" ~ 2,
    coping_strategy == "crisis" ~ 3,
    coping_strategy == "emergency" ~ 4
  ))

```



Plotting coping strategies

```{r}
ggplot(main_ds, aes(x = O_1_respondent_consent, fill = coping_strategy))+
  geom_bar(position = "fill", stat = "count", width = 0.5)+
  theme_minimal()+
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank()
  )
```
Use of coping strategies frequencies

```{r}
main_ds |> select(coping_strategy) |> tbl_summary()

main_ds |> select(oblast_total, coping_strategy) |> tbl_summary(by = oblast_total) |> add_overall()
```



ECMEN

Preparation:

Preparing variables, removing text variables (others)

```{r}
# hh size as numeric
main_ds <- main_ds |> mutate(A_8_hh_members = as.numeric(A_8_hh_members))

# get expenditures variables
exp_vars_all <- names(main_ds)[(str_starts(names(main_ds), "I_3_") | str_starts(names(main_ds), "I_4_") | str_starts(names(main_ds), "I_5_")) & !str_detect(names(main_ds), "_calc") & !str_detect(names(main_ds), "_0_")]

exp_vars_all <- exp_vars_all[! exp_vars_all %in% c("I_5_10_all_other_infrequent_expenditures_please_specify_other", "I_4_15_all_other_frequent_expenditures_please_specify_other", "I_3_9_other_cashcredit_other")]

# expenditure variables as numeric
main_ds <- main_ds |> mutate(across(exp_vars_all, as.numeric))
```


Transforming variables per capita

```{r}
for(var in exp_vars_all){
  main_ds <- main_ds |> mutate(!!sym(paste0(var, "_per_cap")):=!!sym(var)/A_8_hh_members)
}

#getting variables names
exp_vars_all_pc <- names(main_ds)[str_ends(names(main_ds), "_per_cap")]
```


Replacing values above 99-th percentile with the value of 99-th percentiile

```{r}
for(var in exp_vars_all_pc){
   x_q99 <- quantile(as.numeric(main_ds[, var]), 0.99, na.rm = T, type = 2)
    print(paste0(var, " ", x_q99))
#   main_ds <- main_ds |> mutate(!!var:=ifelse(!!sym(var) > x_q99, x_q99, !!sym(var)))
 }
```

replacing outliers to NA for further imputation

```{r}
for (v in exp_vars_all_pc){
  med <- median(main_ds[, paste0(v)], na.rm = T)
  sd <- sd(main_ds[, paste0(v)], na.rm = T)
  thr <- med+sd*3
  print(paste0(v, " ", thr))
  main_ds <- main_ds |> mutate(!!v:=ifelse(!!sym(v)>thr, NA, !!sym(v)))
}
```


Checking the amount of missing data in expenditures

```{r}
nas <- as.data.frame(matrix(nrow = 0,ncol = 2))
nas <- setNames(nas, c("NAs", "Varr"))
rownames(nas) <- NULL
for (v in exp_vars_all_pc){
temp <- main_ds |> 
  summarise(sum(is.na(!!sym(v)))) |>
  mutate(name = paste0(v))
 temp <-  setNames(temp, c("NAs", "Var"))
rownames(temp) <- NULL
nas <- rbind(temp, nas)
}
nas <- nas |> 
  mutate(prp_missing = NAs/nrow(main_ds))
nas |> arrange(desc(prp_missing))
```
Note: Even though the largest share of NAs is present in the variables, that are not part of ECMEN indicator, for some ECMEN variables, % of missings is > 10%. Thus, to handle missing values, imputation will be used.


Preparing predictor variables


1. hh size, - numeric
2. hhs with children - binary
3. hohh's age - numeric
 - 4. rural/urban (raion) - categorical (bin)
5. oblast - categorical (bin)
6. Total income - numeric
7. FCS - numeric
8. HoHH's gender
9. HHs with PwDs
10. Own production as food source
11. Assistance received (any kind)

Transforming predictor variables

```{r}
# Extracting HoHH's gender and age data

# 1 - hh size
main_ds$A_8_hh_members <- as.numeric(main_ds$A_8_hh_members)

# 2 - hh with children
children <- hh_members |>
  mutate(child = ifelse(as.numeric(A_14_hh_member_age) < 18, 1, 0))

children <- children |>
  group_by(uuid) |>
  summarise(count_less_than_18 = sum(child))

main_ds <- main_ds |>
  left_join(children)
#main_ds$count_less_than_18 <- as.numeric(main_ds$count_less_than_18)
main_ds$hh_with_children <- ifelse(main_ds$count_less_than_18 > 0, 1, 0)

# 3 resp's age
main_ds$A_1_respondent_age <- as.numeric(main_ds$A_1_respondent_age)

#rural urban
# = to be added

# 5. oblast (to dummy)
main_ds <- main_ds |> 
  mutate(n = 1) |>
  pivot_wider(names_from = oblast_total, values_from = n, values_fill = 0, names_prefix = "obl_")

# 
income_vars <- names(main_ds)[grepl("I_2_", names(main_ds))]
income_vars <- income_vars[!grepl("_0_", income_vars)]
main_ds <- main_ds |> mutate(across(all_of(income_vars), ~as.numeric(.)))
main_ds <- main_ds |> 
  mutate(total_income_calc=rowSums(across(all_of(income_vars)), na.rm = T))
main_ds$total_income_calc <- as.numeric(main_ds$total_income_calc)
main_ds$total_income_pc <- main_ds$total_income_calc/as.numeric(main_ds$A_8_hh_members)
m <- median(main_ds$total_income_pc, na.rm = T)
sd <- sd(main_ds$total_income_pc, na.rm = T)

main_ds$total_income_pc <- ifelse(main_ds$total_income_pc > m+sd*3 | main_ds$total_income_pc < m-sd*3, NA, main_ds$total_income_pc)

# respondents's gender

main_ds <- main_ds |> 
  mutate(n = 1) |>
  pivot_wider(names_from = A_2_respondent_sex, values_from = n, values_fill = 0, names_prefix = "resp_")

# Employed members in HH
employed <- hh_members |> 
  mutate(n = 1) |>
  pivot_wider(everything(), names_from = A_17_hh_member_employment, values_from = n, values_fill = 0)
  
employed <- employed |>
mutate(count_employed = rowSums(across(all_of(c("in_paid_work_permanent_job", "in_paid_work_temporary_job", "student_and_in_paid_work", "unofficially_employed_informal_work", "self_employed_owns_business", "in_military_service", "in_paid_work_daily_labour")))))

employed <- employed |>
  group_by(uuid) |>
  summarise(count_employed = sum(count_employed))

main_ds <- main_ds |>
  left_join(employed)


main_ds <- 
  main_ds |>
  mutate(hh_with_employed = ifelse(as.numeric(count_employed)>0, 1, 0))

# PwDs in HH
hh_members <- hh_members |>
  
  mutate(WG_disability3 = case_when(
    A_19_wg_vis_ss %in% c("a_lot_of_difficulty", "cannot_do_at_all") | 
      A_20_wg_hear_ss %in% c("a_lot_of_difficulty", "cannot_do_at_all") | 
      A_21_wg_mob_ss %in% c("a_lot_of_difficulty", "cannot_do_at_all") | 
      A_22_wg_cog_ss %in% c("a_lot_of_difficulty", "cannot_do_at_all") | 
      A_23_wg_sc_ss %in% c("a_lot_of_difficulty", "cannot_do_at_all") | 
      A_24_wg_com_ss %in% c("a_lot_of_difficulty", "cannot_do_at_all") ~ 1,
    
    A_19_wg_vis_ss %in% c("dont_know", "prefer_not_to_answer") | 
      A_20_wg_hear_ss %in% c("dont_know", "prefer_not_to_answer") | 
      A_21_wg_mob_ss %in% c("dont_know", "prefer_not_to_answer") | 
      A_22_wg_cog_ss %in% c("dont_know", "prefer_not_to_answer") | 
      A_23_wg_sc_ss %in% c("dont_know", "prefer_not_to_answer") | 
      A_24_wg_com_ss %in% c("dont_know", "prefer_not_to_answer") ~ NA_real_,
    
    is.na(A_19_wg_vis_ss) | 
      is.na(A_20_wg_hear_ss) | 
      is.na(A_21_wg_mob_ss) | 
      is.na(A_22_wg_cog_ss) | 
      is.na(A_23_wg_sc_ss) | 
      is.na(A_24_wg_com_ss) ~ NA_real_,
    
    TRUE ~ 0
  ))

pwd <- hh_members |>
  group_by(uuid) |> 
  summarise(count_disabled = sum(WG_disability3))

main_ds <- main_ds |>
  left_join(pwd)

main_ds <- main_ds |> mutate(hhs_with_pwd = ifelse(as.numeric(count_disabled)>0, 1, 0))

# fcs

# own production as a food source

main_ds$own_production <- as.numeric(main_ds$`G_4_main_food_sources/own_production`)

# assistance received

main_ds <- main_ds |>
  mutate(L_4_assistance_received_factor = as.factor(L_4_assistance_received))
```

Imputing missing expenditures data using MICE multiple imputations

1. selecting variables of interest (with expected influence on expenditures)

```{r}
predictors <- c("total_income_pc", "obl_UA32", "obl_UA26", "obl_UA48", "obl_UA46", "obl_UA59", "obl_UA63", "obl_UA56", "obl_UA68", "obl_UA61", "obl_UA05", "obl_UA73", "obl_UA21", "obl_UA12", "obl_UA18", "obl_UA51", "obl_UA35", "obl_UA07", "obl_UA80", "obl_UA74", "obl_UA71", "obl_UA23", "obl_UA53", "obl_UA65", "obl_UA14", "A_1_respondent_age", "hh_with_children", "A_8_hh_members", "resp_male", "resp_female", "hh_with_employed", "own_production", "fcs", "L_4_assistance_received_factor")

ds_for_imp <- main_ds |>
  select(all_of(c(predictors, exp_vars_all_pc)))
```

2. Imputing data (MICE imputation; "predictive means matching" method), 5 datasets

```{r}
imp <- mice(ds_for_imp, m = 5, maxit = 10, method = "pmm", seed = 123)
impdat <- complete(imp,action="long",include = T)
```

Plotting imputed datasets for exploration

```{r}
plot_list <- list()

for(i in 1:length(exp_vars_all_pc)){
  p <- ggplot(impdat[,], aes(y = !!sym(exp_vars_all_pc[i]), x = .id, color = as.factor(.imp)))+
  geom_line()
  plot_list[[i]] <- p
}

#pdf("plots_imp.pdf", onefile = T)
for (i in 1:length(exp_vars_all_pc)) {
    print(plot_list[[i]])
}
#dev.off()
```
Checking the central tendency of imputed data

```{r}
for(v in exp_vars_all_pc){
  print(with(impdat, by(impdat, .imp, function(x) c(mean(x[,paste0(v)] , na.rm = T),sd(x[,paste0(v)], na.rm = T), median(x[, paste0(v)], na.rm = T)))))
  print(v)
}
```
Even thought by multiple imputation methodology it is strongly adviced to perform analysis on each imputed dataset, because of the composite nature of the indicator, one dataset is to be selected. The dataset 1 is to be selected (as the one, that fits the original distribution ( visual analysis of plots + central tendencies))

```{r}
good_im_data <- impdat[impdat$.imp == 1,]
names(good_im_data) <- paste0("imp1_", names(good_im_data))
good_im_data_exp <- good_im_data |> select(starts_with("imp1_I_"))

main_ds <- cbind(main_ds, good_im_data_exp)
```


Computing food expenditures (detailed)

```{r}
exp_food_vars <- names(main_ds)[str_starts(names(main_ds), "imp1_I_3_") & !str_ends(names(main_ds), "_calc") & str_ends(names(main_ds), "_per_cap")]

exp_food_cashcredit_vars <- exp_food_vars[str_detect(exp_food_vars, "_cashcredit")]
exp_food_own_vars <- exp_food_vars[str_detect(exp_food_vars, "own_produced")]

main_ds <- main_ds |> mutate(exp_food_1m_cashcredit = rowSums(across(all_of(exp_food_cashcredit_vars), as.numeric), na.rm = T))

main_ds <- main_ds |> mutate(exp_food_1m_own_produced = rowSums(across(all_of(exp_food_own_vars), as.numeric), na.rm = T))

# 2023
main_ds <- main_ds |> mutate(exp_food_total_ii = exp_food_1m_cashcredit+exp_food_1m_own_produced)

## UPD fix

#main_ds <- main_ds |> mutate(exp_food_1m_cashcredit = ifelse(Organization == "WFP", (exp_food_7d_cashcredit/7)*30, exp_food_7d_cashcredit),
#                             exp_food_1m_own_produced = ifelse(Organization == "WFP",(exp_food_7d_own_produced/7)*30, exp_food_7d_own_produced))

#main_ds <- main_ds |> mutate(exp_food_1m_total_aggr = exp_food_1m_cashcredit+exp_food_1m_own_produced)
```

Calculating NF aggregates

```{r}
## NF with 1m recall period
exp_nfi_short_vars <- names(main_ds)[str_starts(names(main_ds), "imp1_I_4_") & str_ends(names(main_ds), "_per_cap")]
exp_nfi_short_vars <- exp_nfi_short_vars[!exp_nfi_short_vars %in% c( "imp1_I_4_12_productive_assets_any_items_used_to_help_generate_income_ex_sewing_machine_tools_per_cap",                                                 #"imp1_I_4_15_all_other_frequent_expenditures_please_specify_per_cap", 
                                                                     "imp1_I_4_1_rent_per_cap", "imp1_I_4_15_all_other_frequent_expenditures_please_specify_other_per_cap"
                                                                     #"imp1_I_4_9_recreation_sport_culture_per_cap"
                                                                     )]

main_ds <- main_ds |> mutate(exp_nfi_short_1m = rowSums(across(all_of(exp_nfi_short_vars), as.numeric), na.rm = T))

## NF with 6m recall period

exp_nfi_long_vars <- names(main_ds)[str_starts(names(main_ds), "imp1_I_5_") & str_ends(names(main_ds), "_per_cap")]
exp_nfi_long_vars <- exp_nfi_long_vars[!exp_nfi_long_vars %in% c("imp1_I_5_8_remittances_per_cap",
#"imp1_I_5_10_all_other_infrequent_expenditures_please_specify_per_cap",
"imp1_I_5_9_savings_ie_money_set_aside_for_the_longterm_per_cap", "imp1_I_5_10_all_other_infrequent_expenditures_please_specify_other_per_cap",
"imp1_I_5_7_debt_repayment_per_cap")]

main_ds <- main_ds |> mutate(exp_nfi_long_6m = rowSums(across(all_of(exp_nfi_long_vars), as.numeric), na.rm = T))

main_ds <- main_ds |> mutate(exp_nfi_long_1m = exp_nfi_long_6m/6)

main_ds <- main_ds |> mutate(exp_nfi_total = exp_nfi_short_1m + exp_nfi_long_1m)
```


Dealing with unrealistically low values (Food expenditures)

```{r}
main_ds |> count(exp_food_total_ii < 500)

  main_ds <- main_ds |> mutate(exp_food_total_ii = ifelse(exp_food_total_ii < 500, 500, exp_food_total_ii))
  
  
```
Truncanting outliers (replacing values above 99% with the value of 99%)

```{r}

q99f <- quantile(main_ds$exp_food_total_ii, probs = 0.99)
q99f

main_ds <- main_ds |> mutate(exp_food_total_ii = ifelse(exp_food_total_ii > q99f, q99f, exp_food_total_ii))
  
```


Imputing 0-s in NF expenditures with medians by raion

```{r}
main_ds <- main_ds |> mutate(raion_total = coalesce(raion, O_5_current_location_raion))

main_ds <- main_ds |> group_by(raion_total) |> mutate(exp_nfi_total_ii = ifelse(exp_nfi_total == 0 | is.na(exp_nfi_total), median(exp_nfi_total), exp_nfi_total)) |> ungroup()

#2023
# hist(main_ds$exp_nfi_total)
# main_ds |> count(exp_nfi_total == 0)
# main_ds <- main_ds |> mutate(exp_nfi_total_ii = exp_nfi_total)
```

Truncanting outliers (replacing values above 99% with the value of 99%)

```{r}
summary(main_ds$exp_nfi_total_ii)
q99nf <- quantile(main_ds$exp_nfi_total_ii, probs = 0.99)
q99nf

main_ds <- main_ds |> mutate(exp_nfi_total_ii = ifelse(exp_nfi_total_ii > q99nf, q99nf, exp_nfi_total_ii))
  
```

Total expenditures
  
```{r}
main_ds <- main_ds |> mutate(exp_total = exp_food_total_ii+exp_nfi_total_ii)

#main_ds <- main_ds |> mutate(exp_total = exp_food_1m_cashcredit+exp_nfi_total)
```

Removing cash assistance

1. Aggregating it per capita and removing outliers
```{r}
main_ds <-  main_ds |> mutate(ca_pc  = as.numeric(I_2_10_ngo_or_charity_assistance_)/A_8_hh_members)

ca99 <- quantile(main_ds$ca_pc, 0.99, na.rm = T, type = 6)
main_ds <- main_ds |> mutate(ca_pc = ifelse(ca_pc > ca99, ca99, ca_pc))

main_ds <- main_ds |> mutate(ca_receivers = ifelse(is.na(ca_pc) | ca_pc == 0, "Not CA receiver", "CA receiver"))

```

2. Removing non-consumption part from CA

```{r}
main_ds <- main_ds |> mutate(non_consumption = I_4_12_productive_assets_any_items_used_to_help_generate_income_ex_sewing_machine_tools_per_cap+I_4_9_recreation_sport_culture+(I_5_7_debt_repayment_per_cap+I_5_8_remittances_per_cap+I_5_9_savings_ie_money_set_aside_for_the_longterm_per_cap)/6)

main_ds <- main_ds |> mutate(
  exp_cons_non_cons = non_consumption+exp_total,
  non_cons_ratio = non_consumption/exp_cons_non_cons)

mean(main_ds$non_cons_ratio, na.rm = T)

main_ds <- main_ds |> group_by(ca_receivers) |> mutate(ca_pc_adjusted = ca_pc*(1-mean(non_cons_ratio, na.rm = T))) |> ungroup()
```

Substracting cash assistance and truncationg negative values to 0

```{r}
#main_ds <- main_ds |> mutate(ca_pc_adjusted = ifelse(is.na(ca_pc_adjusted), 0, ca_pc_adjusted))
#main_ds <- main_ds |> mutate(exp_total_pc_no_ca = exp_total - ca_pc_adjusted)
#main_ds <- main_ds |> mutate(exp_total_pc_no_ca = ifelse(exp_total_pc_no_ca < 0, 0, exp_total_pc_no_ca))
# 2023

main_ds <- main_ds |> mutate(ca_pc = ifelse(is.na(ca_pc),0, ca_pc),
                             exp_total_pc_no_ca = exp_total - ca_pc)
main_ds <- main_ds |> mutate(exp_total_pc_no_ca = ifelse(exp_total_pc_no_ca < 0, 0, exp_total_pc_no_ca))
```

MEB and SMEB thresholds

```{r}
#meb_gvt_old <- 5865
#smeb_gvt_old <- 2589

#meb_gvt_infl
#smeb_gvt_infl

#meb_cwg_1 <- 6318
#smeb_cwg_1 <- 2691

meb_cwg_2 <- 6318
smeb_cwg_2 <- 3000

#meb_med <- median(main_ds$exp_total_pc_no_ca)
#smeb_med <- meb_med/2


```

Calculating economic capacity variables

```{r}
  meb <- meb_cwg_2
  smeb <- smeb_cwg_2
  main_ds <- main_ds |> mutate(ECMEN_above_MEB = ifelse(exp_total_pc_no_ca > meb, 1, 0),
                             ECMEN_above_smeb = ifelse(exp_total_pc_no_ca > smeb, 1, 0),
                             ECMEN_text = case_when(
                               exp_total_pc_no_ca > meb ~ "Above MEB",
                               exp_total_pc_no_ca <= meb & exp_total_pc_no_ca >= smeb ~ "Between MEB and SMEB",
                               exp_total_pc_no_ca < smeb ~ "Below SMEB"
                             ),
                             ec_vulnerability = case_when(
                               exp_total_pc_no_ca > meb ~ 1,
                               exp_total_pc_no_ca <= meb & exp_total_pc_no_ca >= smeb ~ 3,
                               exp_total_pc_no_ca < smeb ~ 4
                             ))



```

Re-mutating oblast variable

```{r}
main_ds <- main_ds |> mutate(oblast_total = coalesce(oblast, O_4_current_location_oblast))
```



```{r}
main_ds |> select(ec_vulnerability, oblast_total) |> tbl_summary(by = oblast_total) |> add_overall()

```

Aggregating coping capacity

```{r}

    main_ds <- main_ds |> mutate(coping_capacity = (lcs_cc + ec_vulnerability)/2)

```

Classifying per CARI scale

```{r}

  
  main_ds <- main_ds |> 
    mutate(cari_score= (CARI_current_status + coping_capacity)/2)

  
main_ds <- main_ds |> mutate(cari_lvl = case_when(
  cari_score <= 1.25 ~ "Food secure",
  cari_score >= 1.5 & cari_score <= 2.25 ~ "Marginally food secure",
  cari_score >= 2.5 & cari_score <= 3.25 ~ "Moderately food insecure",
  cari_score >= 3.5 ~ "Severely food insecure"))

```

Plotting CARI levels

```{r}
ggplot(main_ds, aes(x = O_1_respondent_consent, fill = cari_lvl))+
  geom_bar(position = "fill", stat = "count", width = 0.5)+
  theme_minimal()+
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank()
  )
```
CARI levels frequencies

```{r}
#prop.table(table(main_ds$cari_lvl))

main_ds |> select(starts_with("cari_lvl"), oblast_total) |> tbl_summary(by = oblast_total) |> add_overall()
  
```
CARI table

```{r}
main_ds |> select(cari_lvl, CARI_current_status, ec_vulnerability, coping_strategy) |> tbl_summary(by = cari_lvl) |> add_overall()
  
  
  #group_by(cari_lvl, CARI_current_status, ec_vulnerability, coping_strategy) |> summarise(n = n()) |> mutate(prop = n/sum(n)*100) |> pivot_wider(names_from = cari_lvl, values_from = c(prop))

```


Write the dataset with calculated variables

```{r}
write.xlsx(main_ds, "MSNA23_clean_with_CARI_230829_1.xlsx")
```

